---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Load required libraries
```{r echo = TRUE}
library(dplyr, warn.conflicts = FALSE, quietly = TRUE)
library(knitr)
library(lattice)
```

## Loading and preprocessing the data 
### Load the data
```{r echo = TRUE, results='asis'}
if (!exists("activity.csv") & exists("activity.zip")){
  unzip("activity.zip")
}
activity_raw <- read.csv("activity.csv")
```

### Process/transform the data (if necessary) into a format suitable for your analysis
```{r echo = TRUE, results='asis'}
# Convert the character type dates into Date class
activity_raw$date <- as.Date(activity_raw$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?
### Calculate the total number of steps taken per day
```{r echo = TRUE}
total_number_of_steps_per_day <- activity_raw %>% 
  group_by(date) %>% 
  summarise(total.steps = sum(steps, na.rm = TRUE))

head(total_number_of_steps_per_day)
```

### Make a histogram of the total number of steps taken each day
```{r echo = TRUE, results = 'asis'}
with(total_number_of_steps_per_day, hist(total.steps, 
                                         xlab = c("Total Steps per day"),
                                         main = c("Histogram of total steps per day"),
                                         col = "skyblue3",
                                         breaks = 8))
```

### Calculate and report the mean of the total number of steps taken per day
```{r echo = TRUE}
sprintf("Mean Total Steps Per Day = %0.0f",
        total_number_of_steps_per_day %>% summarise(mean(total.steps, na.rm = TRUE)))
```

### Calculate and report the median of the total number of steps taken per day
```{r echo = TRUE}
sprintf("Median Total Steps Per Day = %0.0f",
        total_number_of_steps_per_day %>% summarise(median(total.steps, na.rm = TRUE)))
```

## What is the average daily activity pattern?
### Make a time series plot ( i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r echo = TRUE, results = 'asis'}
mean_daily_activity <- activity_raw %>% group_by(interval) %>% summarise(mean.steps = mean(steps, na.rm = TRUE))

with(mean_daily_activity, plot(interval, mean.steps, 
                               type='l',
                               xlab = "Intervals",
                               ylab = "Average Steps taken",
                               main = "Daily Activity Pattern"))
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r echo = TRUE, results = 'asis'}
sprintf("5-minute interval with maximum number of steps is %0.0f",mean_daily_activity$interval[which.max(mean_daily_activity$mean.steps)])
```

## Imputing missing values
### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs
```{r echo = TRUE, results = 'asis'}
sprintf("Total number of NAs = %0.0f", sum(is.na(activity_raw$steps)))
```

### Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

### Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r echo = TRUE, results = 'asis'}

# Find the rows with NA steps
indices_with_NA <- which(is.na(activity_raw$steps))

# Make a copy of the activity dataframe that will be imputed
activity <- activity_raw

# For each Na entry in the steps column, replace it with the mean across days for the corresponding interval
for (current_index in indices_with_NA)
{
  activity$steps[current_index] <- mean_daily_activity$mean.steps[mean_daily_activity$interval == activity$interval[current_index]]
}

```

### Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? 
```{r echo = TRUE, results = 'asis'}

# Calculate total number of steps taken each day
total_number_of_steps_per_day_imputed <- activity %>% 
  group_by(date) %>% 
  summarise(total.steps = sum(steps, na.rm = TRUE))

# Create a histogram
with(total_number_of_steps_per_day_imputed, hist(total.steps, 
                                         xlab = c("Total Steps per day"),
                                         main = c("Histogram of total steps per day after imputing missing values"),
                                         col = "skyblue3",
                                         breaks = 8))

sprintf("Mean Total Steps Per Day after imputing missing values = %0.0f",
        total_number_of_steps_per_day_imputed %>% summarise(mean(total.steps, na.rm = TRUE)))
sprintf("Median Total Steps Per Day after imputing missing values = %0.0f",
        total_number_of_steps_per_day_imputed %>% summarise(median(total.steps, na.rm = TRUE)))
```

### What is the impact of imputing missing data on the estimates of the total daily number of steps?
Imputing the missing values has changed the distribution of the steps across days. The imputing strategy of replacing the missing values with non-zero mean values has increased both the mean and median of the total number of steps across days 

## Are there differences in activity patterns between weekdays and weekends?
### Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r echo = TRUE, results = 'asis'}
activity <- mutate(activity, type.of.day = as.POSIXlt(activity$date)$wday)
activity$type.of.day[which(activity$type.of.day < 6 & activity$type.of.day > 0)] <- c("weekday")
activity$type.of.day[which(activity$type.of.day == 0 | activity$type.of.day == 6)] <- c("weekend")
activity$type.of.day <- as.factor(activity$type.of.day)
```

### Make a panel plot containing a time series plot (i.e.type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r echo = TRUE, results = 'asis'}
# Calculate the average total number of steps across intervals by weekday or weekend 
activity_type_of_day <- activity %>% group_by(type.of.day, interval) %>% summarise(mean.steps = mean(steps, na.rm = TRUE))

# Generate the panel plot
xyplot(mean.steps ~ interval | type.of.day,
    data = activity_type_of_day, type = c("l"),
    layout = c(1,2),
    xlab = "Interval",
    ylab = "Number of Steps")
```